\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-27 paper.tex
Clarify espalier

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The figure shows the pedigree of eight diploid individuals
and their sixteen constituent genomes (each consisting of a single chromosome),
along with paths of genetic inheritance.
Here, and throughout,
nodes are labelled with lowercase alphabetical letters
rather than integer identifiers, to avoid confusion with genomic intervals.
Thus individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the two genomes belonging to an individual parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations (A), the corresponding gARG (B), and finally the corresponding
sequence of local trees along the
genome (C).
% Jerome: this feels unecessary and "genealogy" is often interpreted as
% "pedigree" so needs to be used very carefully.
% The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
% which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes;
further discussion on
how local trees are embedded in an ARG is provided in Section~\ref{sec-ARG-and-local-trees}
% Following the above definitions, corresponding data encoding for
% this gARG is shown in Tab.~\ref{tab-gARG-data}.

@@

\section{ARGs and local trees}
\label{sec-ARG-and-local-trees}
% (We use ``local'' rather than
% ``marginal''
% % could list loads here, but not much point.
% trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
% to avoid the potential confusion with the
% statistical concept of marginal distributions.)
The relationship between an ARG and its corresponding
local trees is subtle and important.
A fundamental property of genetics is that a
given DNA nucleotide is inherited from exactly one parent genome,
both at an organismal and cell-by-cell level
(Appendix~\ref{sec-cell-lineages-and-args}).
By definition, these chains of single-parent inheritance give rise,
over time, to a tree structure.
As a result of recombination, adjacent nucleotides can have
different paths of inheritance, and
an ARG encodes the entire ensemble of local trees along the
% Said "sample of genomes" originally, but using term "genome" in two
% different ways in the same sentence then.
genome for a given set of sample nodes.
% This is for Yan - to me this is a given, but maybe some people feel
% the following is overly formal?
Precisely defining the process by which local trees are extracted
from an ARG is essential to our understanding of how ARGs and local trees
are related, and we require a concrete mathematical structure
to describe the local trees.

@@

This is not the prevailing view, however.
\cite{kuhner2017consensus} argue that the
``interval-tree'' representation of
an ARG (the local trees and the genome intervals they cover)
``does not contain all of the information in the underlying ARG: it lacks the
number of recombinations occuring at each site, the times at which
recombinations occurred, and the specific sequences involved as recombination
partners.''
\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG...~contains more information than the series of tree
sequences along the genome''.
% Could cite some more? E.g. KwARG paper says
% "We note that ARGs contain more information than local trees"
% If even Ana and Jotun are wrong about this it then it must be
% significant! ;)
These statements that an ARG
contains more information than its local trees are only true
if we use a representation of the local trees that discards
information about those local trees that can be derived from the ARG.

There are two properties of how evolutionary trees are conventionally
represented that lead to this
% We can rephrase... suggestions?
disagreement about the relationship between local trees and an ARG.
Firstly, the internal nodes of evolutionary trees are usually
considered to be \emph{unlabelled}, or equivalently, labelled by the leaves which they subtend.
The same canonical labelling cannot be used for internal ARG nodes because the leaves they subtend will typically vary by genomic position. If we do not label the tree nodes in a way which is persistent across the sequence of local trees in the ARG, we lose the fact that
the \emph{same} ancestors sometimes persist across multiple trees.
Defining ARG nodes as integers and using the oriented tree encoding
explicitly labels internal nodes, and makes the relationship between
tree and ARG nodes clear.
The second property of how evolutionary trees
are conventionally represented that is unhelpful in the context of ARGs is their
focus on branching points (coalescences), i.e.\ nodes that have two or more children.
As the introductory paragraph of this section emphasised,
parent-child relationships are what fundamentally define a tree,
and branching points can be seen as incidental. This is reflected
by the oriented tree encoding which simply stores the local
parent-child relationships, and does not, for example,
directly tell us how many children a particular node has.
The local tree at a given position records the \emph{path} through
the ARG; if this path omits nodes that are not
branching points in the local tree (such as \noderef{e} in Fig.~\ref{fig-arg-in-pedigree}),
we lose information about the ARG.
We return to this point in the following two sections,
when we discuss ``locally unary'' nodes and the simplification process.

It is important that we make the distinction here between the local
trees that we can derive from a known ARG (as just discussed),
and an ARG that we can derive from a sequence of \emph{estimated}
local trees.
The ARG inference method
\espalier~\citep{rasmussen2022espalier} is illustrative in this context.
It begins by splitting an input sequence alignment into
segments that are assumed to be non-recombining. Within
each segment, an initial local tree is estimated using
standard phylogenetic methods. By necessity, these local
trees will contain internal nodes that are unlabelled and
consist only of branching points: there is no information
shared between the independent tree estimation steps
across segments. Part of the task of stitching
these trees together
into an ARG is then, essentially, to generate labels for
the internal nodes, and decide which nodes persist
across multiple local trees.
% TODO add a sentence here saying how it does this.
% that's what we're interested in illustrating here.
\espalier\ is specifically interested
in the details of recombination events,
and also attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination between these partially reconciled local trees.
A similar problem is tackled by \arbores~\citep{heine2018bridging}, which explores a space of ARGs by fixing a starting ARG, and then repeatedly erasing and resimulating candidate SPR moves connecting two fixed, leaf-labelled local trees.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.
% I don't really know if this is true, it's just a way of wrapping the
% section up by saying that other things we thought were settled might
% be worth thinking about again.
The combinatorial formulation of ARGs and local trees provided here
may help clarify these fundamental questions.

>>>>

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The figure shows the pedigree of eight diploid individuals
and their sixteen constituent genomes (each consisting of a single chromosome),
along with paths of genetic inheritance.
Here, and throughout,
nodes are labelled with lowercase alphabetical letters
rather than integer identifiers, to avoid confusion with genomic intervals.
Thus individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the two genomes belonging to an individual parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations (A), the corresponding gARG (B), and finally the corresponding
sequence of local trees along the
genome (C).
% Jerome: this feels unecessary and "genealogy" is often interpreted as
% "pedigree" so needs to be used very carefully.
% The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
% which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes;
further discussion on
how local trees are embedded in an ARG is provided in Section~\ref{sec-ARG-and-local-trees}.
% Following the above definitions, corresponding data encoding for
% this gARG is shown in Tab.~\ref{tab-gARG-data}.

@@

\section{ARGs and local trees}
\label{sec-ARG-and-local-trees}
% (We use ``local'' rather than
% ``marginal''
% % could list loads here, but not much point.
% trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping,kelleher2016efficient}
% to avoid the potential confusion with the
% statistical concept of marginal distributions.)
The relationship between an ARG and its corresponding
local trees is subtle and important.
A fundamental property of genetics is that a
given DNA nucleotide is inherited from exactly one parent genome,
both at an organismal and cell-by-cell level
(Appendix~\ref{sec-cell-lineages-and-args}).
These paths of single-parent inheritance give rise,
by definition, to a tree structure.
As a result of recombination, adjacent nucleotides can have
different paths of inheritance, and
an ARG encodes the entire ensemble of local trees along the
% Said "sample of genomes" originally, but using term "genome" in two
% different ways in the same sentence then.
genome for a given set of sample nodes.
% This is for Yan - to me this is a given, but maybe some people feel
% the following is overly formal?
Precisely defining the process by which local trees are extracted
from an ARG is essential to our understanding of how ARGs and local trees
are related, and we require a concrete mathematical structure
to describe the local trees.

@@

This is not the prevailing view, however.
\cite{kuhner2017consensus} argue that the
``interval-tree'' representation of
an ARG (the local trees and the genome intervals they cover)
``does not contain all of the information in the underlying ARG: it lacks the
number of recombinations occuring at each site, the times at which
recombinations occurred, and the specific sequences involved as recombination
partners.''
\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG...~contains more information than the series of tree
sequences along the genome''.
% Could cite some more? E.g. KwARG paper says
% "We note that ARGs contain more information than local trees"
% If even Ana and Jotun are wrong about this it then it must be
% significant! ;)
These
% and similar, \citep[e.g.][]{}
statements that an ARG
contains more information than its local trees are true
if we represent local trees in their conventional form,
but this form discards important information that is available from the ARG.

There are two properties of how evolutionary trees are conventionally
represented that lead to this
disagreement about the relationship between local trees and an ARG.
Firstly, the internal nodes of evolutionary trees are usually considered to be
\emph{unlabelled}, or equivalently, labelled by the leaves which they subtend.
The same canonical labelling cannot be used for internal ARG nodes because the
leaves they subtend will typically vary by genomic position. If we do not label
the tree nodes in a way which is persistent across the sequence of local trees
in the ARG, we lose the fact that the \emph{same} ancestors sometimes persist
across multiple trees.
Defining ARG nodes as integers and using the oriented
tree encoding explicitly labels internal nodes, and makes the relationship
between tree and ARG nodes clear and precise.

% Note: good to keep to paras here to logically separate the two important
% points.
The second property of how evolutionary trees
are conventionally represented that is unhelpful in the context of ARGs is their
focus on branching points (coalescences), i.e.\ nodes that have two or more children.
As the introductory paragraph of this section emphasised,
parent-child relationships are what fundamentally define a tree,
and branching points can be seen as incidental. This is reflected
by the oriented tree encoding which simply stores the local
parent-child relationships, and does not, for example,
directly tell us how many children a particular node has.
The local tree at a given position records the \emph{path} through
the ARG; if this path omits nodes that are not
branching points (such as \noderef{e} in Fig.~\ref{fig-arg-in-pedigree}),
we lose information about the ARG.
We return to this point in the following two sections,
when we discuss ``locally unary'' nodes and the simplification process.

It is important that we make the distinction here between the local
trees that we can derive from a known ARG (as just discussed),
and an ARG that we can derive from a sequence of \emph{estimated}
local trees.
The ARG inference method
\espalier~\citep{rasmussen2022espalier} is illustrative in this context.
It begins by splitting an input sequence alignment into
segments that are assumed to be non-recombining. Within
each segment, an initial local tree is estimated using
standard phylogenetic methods. By necessity, these local
trees will contain internal nodes that are unlabelled and
consist only of branching points: there is no information
shared between the independent tree estimation steps
across segments. Part of the task of stitching
these trees together
into an ARG is then, essentially, to generate labels for
the internal nodes, and decide which nodes persist
across multiple local trees.
\espalier\ approaches this task
by identifying maximimal subtrees
that do not change between pairs of adjacent
local trees and then
heuristically exploring the space of possible
rearrangements of these subtrees.
To derive precise details about recombination events,
\espalier\ then attempts to infer
the precise subtree prune-and-regraft (SPR)
operations~\citep{hein1990reconstructing,song2003on,song2006properties}
induced by recombination between these partially reconciled local trees.
% A similar problem is tackled by \arbores~\citep{heine2018bridging}, which
% explores a space of ARGs by fixing a starting ARG, and then repeatedly erasing
% and resimulating candidate SPR moves connecting two fixed, leaf-labelled local
% trees.
Inferring the SPRs between leaf-labelled trees is
NP-hard~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but it is unclear what the complexity is when there
is a degree of internal node sharing between trees.
% I don't really know if this is true, it's just a way of wrapping the
% section up by saying that other things we thought were settled might
% be worth thinking about again.
The combinatorial formulation of ARGs and local trees provided here
may help clarify these fundamental questions.
